---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

<!-- badges: start -->
[![GitHub release (latest by date)](https://img.shields.io/github/v/release/Public-Health-Scotland/slfhelper)](https://github.com/Public-Health-Scotland/slfhelper/releases/latest)
<!-- badges: end -->

# slfhelper

The goal of slfhelper is to provide some easy to use functions which make 
working with the Source Linkage Files as painless and efficient as possible.

## Installation

You can install slfhelper by using the [remotes package](https://remotes.r-lib.org/) in R:

```{r package_install}
install.packages("remotes")
remotes::install_github("Public-Health-Scotland/slfhelper")
```

## Usage

### Read a file

**Note:** Reading a full file is quite slow, we would always recommend doing a column selection to only keep the variables which you need for your analysis. Just doing this will dramatically speed up the read-time.

```{r read_files}
library(slfhelper)

# Read a full file
ep_1718 <- read_slf_episode("1718")
indiv_1718 <- read_slf_individual("1718")

# Read multiple years
# This will use dplyr::bind_rows() and return the files added together as a single tibble
episode_data <- read_slf_episode(c("1516", "1617", "1718", "1819"))

# Read certain variables 
# It's much faster to choose variables like this
indiv_1718 <- read_slf_individual("1718", columns = c("anon_chi", "hri_scot"))

# Read only data for a certain partnership (HSCP_2018 code)
# This can be a single partnership or multiple by supplying a vector e.g. c(...)
indiv_1718 <- read_slf_individual("1718", partnerships = "S37000001") # Aberdeen City

# Read only data for a certain recid
# This can be a single recid or multiple by supplying a vector e.g. c(...)
ep_1718 <- read_slf_episode("1718", recid = c("01B", "GLS"))
```

All of the above options for reading files can be combined if required. 

There are a few data files included in the package which should be useful.

```{r helper_data}
# Get a list of the variables in a file
slfhelper::ep_file_vars
slfhelper::indiv_file_vars

# See a lookup of Partnership names to HSCP_2018 codes

hscp_lookup <- slfhelper::partnerships
```

### Match on CHI numbers to Anon_CHI (or vice versa)

```{r chi_matching}
library(slfhelper)

# Add real CHI numbers to a SLF
ep_1718 <- read_slf_episode("1718") %>% get_chi()

# Add anon_chi numbers to a cohort
# Optionaly specifiy the name of the CHI variable
cohort <- cohort %>% get_anon_chi(chi_var = "upi_number")
```



