---
title: "slf-documentation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{slf-documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, include = FALSE}
library(slfhelper)
library(tidyverse)
```

## SLFhelper 
`SLFhelper` contains some easy to use functions designed to make working with the Source Linkage Files (SLFs) as efficient as possible. 

### Filter functions:
- `year` returns financial year of interest. You can also select multiple years using `c("1718", "1819", "1920")`
- `recid` returns recids of interest. Selecting this is beneficial for specific analysis.
- `partnerships` returns partnerships of interest. Selecting certain partnerships will reduce the SLFs size.
- `col_select` returns columns of interest. This is the best way to reduce the SLFs size. 

### Data snippets: 
- `ep_file_vars` returns a list of all variables in the episode files.
- `indiv_file_vars` returns a list of all variables in the individual files.
- `partnerships` returns a list of partnership names (HSCP_2018 codes)
- `recid` returns a list of all recids available in the SLFs.
- `ep_file_bedday_vars` returns a list of all bedday related variables in the SLFs. 
- `ep_file_cost_vars` returns a list of all cost related variables in the SLFs. 

### Anon CHI
- Use the function `get_chi()` to easily switch `anon_chi` to `chi`. 
- Use the function `get_anon_chi()` to easily switch `chi` to `anon_chi`. 


### Memory usage in SLFS 

While working with the Source Linkage Files (SLFs), it is recommended to use the features of the SLFhelper package to maximase the memory usage in posit, see [PHS Data Science Knowledge Base](https://public-health-scotland.github.io/knowledge-base/docs/Posit%20Infrastructure?doc=Memory%20Usage%20in%20SMR01.md) for further guidance on memory usage in posit workbench.

Reading a full SLF file can be time consuming and take up resources on posit workbench. In the episode file there are `251 variables` and around `12 million rows` compared to the individual file where there are `193 variables` and around `6 million rows` in each file. This can be reduced by using available selections in SLFhelper to help reduce the size of the SLFs for analysis and to free up resources in posit workbench. 

The tables below show the memory usage of each full size SLF. 


## Episode File          

| Year          | Memory usage (MB)| 
| ------------- |:----------------:|
| 1718          | 2651.2           |
| 1819          | 3196.5           |
| 1920          | 3145.4           |
| 2021          | 2715.6           | 
| 2122          | 2959.3           |
| 2223          | 2995.1           | 
| 2324          | 1894.5           | 


## Individual File          

| Year          | Memory usage (MB)| 
| ------------- |:----------------:|
| 1718          | 1055.6           |
| 1819          | 1057.8           |
| 1920          | 1070.7           |
| 2021          | 1067.3           | 
| 2122          | 1081.2           |
| 2223          | 1098.7           | 
| 2324          | 775.5            |

