---
title: "slf-documentation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{slf-documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(slfhelper)
```

## SLFhelper

`SLFhelper` contains some easy to use functions designed to make working with the Source Linkage Files (SLFs) as efficient as possible.

### Filter functions:

-   `year` returns financial year of interest. You can also select multiple years using `c("1718", "1819", "1920")`
-   `recid` returns recids of interest. Selecting this is beneficial for specific analysis.
-   `partnerships` returns partnerships of interest. Selecting certain partnerships will reduce the SLFs size.
-   `col_select` returns columns of interest. This is the best way to reduce the SLFs size.

### Data snippets:

-   `ep_file_vars` returns a list of all variables in the episode files.
-   `indiv_file_vars` returns a list of all variables in the individual files.
-   `partnerships` returns a list of partnership names (HSCP_2018 codes)
-   `recid` returns a list of all recids available in the SLFs.
-   `ep_file_bedday_vars` returns a list of all bedday related variables in the SLFs.
-   `ep_file_cost_vars` returns a list of all cost related variables in the SLFs.

### Anon CHI

-   Use the function `get_chi()` to easily switch `anon_chi` to `chi`.
-   Use the function `get_anon_chi()` to easily switch `chi` to `anon_chi`.

### Memory usage in SLFS

While working with the Source Linkage Files (SLFs), it is recommended to use the features of the SLFhelper package to maximase the memory usage in posit, see [PHS Data Science Knowledge Base](https://public-health-scotland.github.io/knowledge-base/docs/Posit%20Infrastructure?doc=Memory%20Usage%20in%20SMR01.md) for further guidance on memory usage in posit workbench.

Reading a full SLF file can be time consuming and take up resources on posit workbench. In the episode file there are `r length(slfhelper::ep_file_vars)` variables and around 12 million rows compared to the individual file where there are `r length(slfhelper::indiv_file_vars)` variables and around 6 million rows in each file. This can be reduced by using available selections in SLFhelper to help reduce the size of the SLFs for analysis and to free up resources in posit workbench.

The tables below show the memory usage of each full size SLF.

#### Episode File

| Year | Memory Usage (GiB) |
|------|:------------------:|
| 1718 |         22         |
| 1819 |         22         |
| 1920 |         22         |
| 2021 |         19         |
| 2122 |         21         |
| 2223 |         21         |
| 2324 |         18         |

#### Individual File

| Year | Memory Usage (GiB) |
|------|:------------------:|
| 1718 |        6.8         |
| 1819 |        6.8         |
| 1920 |        7.0         |
| 2021 |        7.0         |
| 2122 |        7.0         |
| 2223 |        7.1         |
| 2324 |        5.1         |

If one can use selection features in SLFhelper, the session memory requirement can be reduced. There are `r length(slfhelper::ep_file_vars)` columns for a year episode file of size around 20 GiB. Hence, on average, a column with all rows takes around 0.1 GiB, which can give a rough estimate on the session memory one needs. Taking Year 1920 as a demonstration, the following tables present various sizes of extracts from the SLF files, from 5 columns to all columns, along with the amount of memory required to work with the data one reads in. Keep in mind that tables below are just recommendations, and that memory usage depends on how one handles data and optimises data pipeline.


#### Episode File
| Column Number | Memory usage (GiB) |           Session Memory Recommendation           |
|---------------|:------------------:|---------------------------------------------------|
| 5             |        0.5         | 4 GiB (4096 MiB)                                  |
| 10            |        1.4         | between 4 GiB (4096 MiB) and 8 GiB (8192 MiB)     |
| 50            |        5.1         | between 8 GiB (8192 MiB) and 16 GiB (16384 MiB)   |
| 150           |         13         | between 20 GiB (20480 MiB) and 38 GiB (38912 MiB) |
| 251           |         22         | between 32 GiB (32768 MiB) and 64 GiB (65536 MiB) |

#### Individual File

| Column Number | Memory usage (GiB) | Session Memory Recommendation |
|---------------|:------------------:|---------------------------------------------------|
| 5             |        0.7         | 4 GiB (4096 MiB)                                  |
| 10            |        0.8         | 4 GiB (4096 MiB)                                  |
| 50            |        2.2         | between 4 GiB (4096 MiB) and 8 GiB (8192 MiB)     |
| 150           |        5.5         | between 8 GiB (8192 MiB) and 16 GiB (16384 MiB)   |
| 193           |        7.0         | between 11 GiB (11264 MiB) and 21 GiB (21504 MiB) |
