---
title: "Using Parquet files with the arrow package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Parquet files with the arrow package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Using Parquet files with the arrow package

The SLFs are available in parquet format. The {arrow} package gives some extra features which can speed up and reduce memory usage even further. You can read only specific columns `read_parquet(file, col_select = c(var1, var2))`.

Using arrow's 'Arrow Table' feature, you can speed up analysis efficiently. To do this, specify `as_data_frame = FALSE` when using SLFhelper and `dplyr::collect()` to read the data.

#### For example:

Imagine a scenario of analysing planned and unplanned beddays in Scotland, there are two ways to read the episode files and do analysis by setting `as_data_frame` to be `TRUE` or `FALSE` as follows.

```{r arrow, eval=FALSE, message=FALSE}
library(slfhelper)

## FAST METHOD
# Filter for year of interest
slf_extract1 <- read_slf_episode(c("1819", "1920"),
  # Select recids of interest
  recids = c("01B", "GLS", "04B"),
  # Select columns
  col_select = c(
    "year", "anon_chi", "recid",
    "yearstay", "age", "cij_pattype"
  ),
  # return an arrow table
  as_data_frame = FALSE
) %>%
  # Filter for non-elective and elective episodes
  dplyr::filter(cij_pattype == "Non-Elective" | cij_pattype == "Elective") %>%
  # Group by year and cij_pattype for analysis
  dplyr::group_by(year, cij_pattype) %>%
  # summarise bedday totals
  dplyr::summarise(beddays = sum(yearstay)) %>%
  # collect the arrow table
  dplyr::collect()

## SLOW and DEFAULT Method
# Filter for year of interest
slf_extract2 <- read_slf_episode(c("1819", "1920"),
  # Select recids of interest
  recids = c("01B", "GLS", "04B"),
  # Select columns
  col_select = c(
    "year", "anon_chi", "recid",
    "yearstay", "age", "cij_pattype"
  ),
  # return an arrow table
  as_data_frame = TRUE # which is default
) %>%
  # Filter for non-elective and elective episodes
  dplyr::filter(cij_pattype == "Non-Elective" | cij_pattype == "Elective") %>%
  # Group by year and cij_pattype for analysis
  dplyr::group_by(year, cij_pattype) %>%
  # summarise bedday totals
  dplyr::summarise(beddays = sum(yearstay))
```

By specifying `as_data_frame = FALSE` when using reading SLF functions, one enjoys great advantages of `parquet` files. One of the advantages is fast query processing by reading only the necessary columns rather than entire rows. The table below demonstrates the huge impact of those advantages.

|                         | Time consumption (seconds) | Memory usage (MiB) |
|-------------------------|:--------------------------:|:------------------:|
| `as_data_frame = TRUE`  |            4.46            |        553         |
| `as_data_frame = FALSE` |            1.82            |        0.43        |

: Comparison of different ways of reading SLF files
