---
title: "using-arrow-table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using-arrow-table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(slfhelper)
```

## Using Parquet files with the arrow package 

The SLFs are available in parquet format. The {arrow} package gives some extra features which can speed up and reduce memory usage even further. You can read only specific columns `read_parquet(file, col_select = c(var1, var2))`. 

Using arrow’s ‘Arrow Table’ feature, you can speed up analysis efficiently. To do this, specify `as_data_frame = FALSE` when using SLFhelper and `dplyr::collect()` to read the data. 


#### For example:

Planned and unplanned beddays in Scotland
```{r chunk1, eval=FALSE, message=FALSE}
# Filter for year of interest
slf_extract <- read_slf_episode(c("1819", "1920"),
  # Select recids of interest
  recids = c("01B", "GLS", "04B"),
  # Select columns
  col_select = c(
    "year", "anon_chi", "recid",
    "yearstay", "age", "cij_pattype"
  ),
  # return an arrow table
  as_data_frame = FALSE
) %>%
  # Filter for non-elective and elective episodes
  filter(cij_pattype == "Non-Elective" | cij_pattype == "Elective") %>%
  # Group by year and cij_pattype for analysis
  group_by(year, cij_pattype) %>%
  # summarise bedday totals
  summarise(beddays = sum(yearstay)) %>%
  # collect the arrow table
  dplyr::collect()
```
